# Normalization for cell-specific biases

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(HDF5Array)
library(scater)
library(scran)
sce <- readRDS("objects/sce.rds")
```

## Setting up genes and cells

We first do some pre-clustering to split the cells into sensible clusters, to avoid violating the non-DE assumption and distorting the size factors.
This is done by applying the `quickCluster` function on the cells in each library.
We do some filtering for high-abundance genes to reduce the effect of zeroes on our rank-based clustering. 

```{r}
by.lib <- split(seq_len(ncol(sce)), sce$Library)
cluster.id <- character(ncol(sce))
for (lib in names(by.lib)) { 
    current <- by.lib[[lib]]
    cur.exprs <- realize(counts(sce)[,current]) # for speed; avoid multiple file reads here.
    ids <- quickCluster(cur.exprs, min.mean=0.1, method="igraph", 
        max.size=3000, irlba.args=list(maxit=1000))
    cluster.id[current] <- paste0(lib, ".", ids)
}
```

## Calculating size factors

We now compute a size factor for each cell using the deconvolution method with the specified parameters.

```{r}
library(scran)
sce <- computeSumFactors(sce, cluster=cluster.id, min.mean=0.1)
summary(sizeFactors(sce))
```

We can have a look at them in more detail, compared to the library size for each cell.

```{r sizefacplot}
plot(sce$total_counts, sizeFactors(sce), log="xy", xlab="Library size", 
    ylab="Size factors", cex=0.2, pch=16)
```

## Normalizing the expression values 

We reset the HDF5 dump file directory and location so that the normalized expression values are retained.

```{r}
new.path <- "objects/norm_exprs.h5"
if (file.exists(new.path)) {
   unlink(new.path)
}
setHDF5DumpFile(new.path)
```

We then calculate normalized log-expression values and save this to a new `HDF5Matrix` object.

```{r}
sce <- normalize(sce)
logcounts(sce)
```

We reset the locations to the defaults for future use.

```{r}
setHDF5DumpFile()
```

<!--
```{r echo=FALSE, results="hide"}
# Saving the new object as well.
saveRDS(file="objects/sce.rds", sce)
```
-->
