# Detecting highly variable genes

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(scran)
sce <- readRDS("objects/sce.rds")
```

## Fitting a trend to the endogenous variances

First, we set up a blocking factor based on the library of origin for each cell.
This ensures that we do not detect variance due to technical differences in sequencing.

```{r}
lib.source <- factor(sce$Library)
table(lib.source)
```

We compute the variance of the normalized log-expression values while blocking on the library of origin for each gene.
We then fit a trend to the variances against the mean of the log-values.
Some fiddling with the trend parameters is necessary to obtain an appropriate fit at high abundances where there is little support.

```{r}
library(scran)
fit <- trendVar(sce, method="loess", parametric=TRUE, 
    use.spikes=FALSE, block=lib.source,
    loess.args=list(span=0.05, control=loess.control(iterations=100)))
```

We then decompose the biological and technical component for each gene.

```{r}
dec <- decomposeVar(fit=fit)
dec <- cbind(rowData(sce)[,1:2], dec)
dec <- dec[order(dec$p.value, -dec$bio),]
head(dec)
```

We examine the mean-variance relationship and verify that the trend was fitted correctly.

```{r hvgplot-tech}
plot(dec$mean, dec$total, pch=16, cex=0.5, xlab="Mean of log-expression", 
    ylab="Variance of log-expression")
curve(fit$trend(x), add=TRUE, col="red")
```

Finally we save the results to file.

```{r}
write.table(file="objects/hvg_output_endog.txt", dec, sep="\t", quote=FALSE, row.names=FALSE)
```

<!--
Also saving the trend.

```{r}
saveRDS(file="objects/trendfit_endog.rds", fit)
```
-->

## Considering a trend for the Poisson noise

An alternative approach is to generate the trend corresponding to Poisson noise.
This can be achieved with the `makeTechTrend` function.
Normally we could just supply the entire `SingleCellExperiment` object to it, but for interests of speed, we'll manually specify the means instead.

```{r}
means <- 2^seq(0.01, max(dec$mean), length.out=100) - 1
pois.trend <- makeTechTrend(means, size.factors=sizeFactors(sce))
```

We replace our endogenous trend with this new trend and then decompose the biological and technical component for each gene.

```{r}
fit2 <- fit
fit2$trend <- pois.trend
dec2 <- decomposeVar(fit=fit2)
dec2 <- cbind(rowData(sce)[,1:2], dec2)
dec2 <- dec2[order(dec2$p.value, -dec2$bio),]
head(dec2)
```

We see that this trend passes below the bulk of variances for endogenous genes.
This is expected for UMI data where the technical noise is much lower and per-gene biological heterogeneity tends to more pronounced.

```{r hvgplot-pois}
plot(dec2$mean, dec2$total, pch=16, cex=0.5, xlab="Mean of log-expression", 
    ylab="Variance of log-expression")
curve(fit2$trend(x), add=TRUE, col="blue")
```

We save the results to file for further use.

```{r}
write.table(file="objects/hvg_output_pois.txt", dec2, sep="\t", quote=FALSE, row.names=FALSE)
```

This trend is probably more appropriate than the endogenous trend for distinguishing between biological and technical variation.
The assumption that most genes are not highly variable is quite strong in a heterogeneous population.

<!--
Also saving the trend.

```{r}
saveRDS(file="objects/trendfit_pois.rds", fit2)
```
-->


